Seismic parameter estimation and the Canadian crust

### 2013-11-11 paper/draftBP.tex
nik edits

The lower energy in the 20-30 km depth window noted in Section \ref{section:discontinuities} for the station profiles given in figure \ref{fig:disc} provides some supporting evidence for seismically distinct upper and lower crustal layers. To better compare results across stations, we normalize each depth profile such that the Moho falls on a common ordinate (Figure \ref{fig:discInterp}, left) and stack the absolute values (Figure \ref{fig:discInterp}, right). The stacked signal has highest amplitude near the surface, where energy is associated with sedimentary layers, and at the Moho. There is a steady and marked decrease in energy over the bottom third of the profile that cannot be ascribed to sidelobe structure associated with the Moho pulse. This observation is ostensibly at odds with the Mooney and Brocher [1987] high reflectivity lower crustal model. Moreover, reflection profiles from {\sc Lithoprobe} transects [Hammer et. al., 2010] frequently display significant reflected energy through the crust with some transects including the Western Superior transect and the Cassiar in the Western Cordillera displaying pronounced pronounced reflectivity within the lower crust. We reconcile the teleseismic and active source observations of transparent versus reflective lower crust by considering the scale of vertical heterogeneity to which the methods are sensitive. Teleseismic body waves include both direct and reflected conversions that can originate from layering with velocity gradients over scale lengths of the order of $\lambda/2$ and $\lambda/4$, respectively, where $\lambda$ is the {\it P}-wavelength [Bostock, 1999]. For 1 Hz teleseismic waves, crustal layering with scale lengths of 1000-2000 m should be resolvable. Higher frequency active source data (10-40 Hz) employing {\it P}-wave sources are sensitive to finer scale vertical structure with scales of 75-300 m. We infer then that the vertical scales of lower crustal reflectivity are limited to scale lengths under 100 m yielding a lower crustal structure that is relatively transparent to teleseismic waves but can yield strongly reflective signatures to near-vertical, active-source seismic profiling.  Some authors, for example Mayer et .al. [1996] in their study of the Rhine Graben, have employed the term Conrad discontinuity to to refer to an interface or transition between transparent upper crust and reflective lower crust. It is possible in this context that the higher energy levels between 15 and 20 km on the 5th principal component in Figure 12, is a manifestation of such
signals, but the lower energy levels ($<5\%$) in this mode, would indicate that Conrad Discontinuity so-defined is at best weakly defined within the Canadian crust. The increased transparency of the lower crust would appear to be better defined and would appear to relate to changes in rheology. Studies on crustal strength [Jackson, 2002] and on brittle and ductile models [Burov and Watts, 2006; Meisner and Rabbel, 2006] as well as the widely noted reduced seismicity of the lower crust indicate changes would support this contention, though see [Deichmann, 1991; Handy and Brun, 2004; Simpson, 1999] for competing evidence.

@@

Seismic parameter estimates for Canada and the Canadian Shield fall, to a first-order approximation, between those of two candidate end-member lithologies: granite geniss and diorite. This intermediate or andesitic composition is in general agreement with other earlier studies of bulk crustal composition [Rudnick and Fountain, 1995; Christensen and Mooney [1995]. The Slave, Superior and Churchill provinces possess comparable seismic parameters implying similar mean crustal compositions, with the Slave province represented as slightly more felsic in average composition. All three provinces cluster about the average value for the Canadian Shield. The Grenville Province, in contrast, displays markedly lower $V_S$ and higher $V_P/V_S$ implying a significant difference in mean composition. The high $V_P/V_S$ estimate, 1.77, is similar to the averaged results from a previous seismic study [Eaton et. al., 2005] that included stations of the central Grenville Orogen, and might be interpreted to indicate that the Grenville province possesses a mean crustal composition corresponding composition that is more mafic than other Shield provinces. This interpretation is consistent with a tectonic model for the the central Grenville Province which suggests that the mid-crust has been completely excised locally [Eaton et. al., 2005; Rivers, 2008]. However, many stations in the south-east of the province also display $V_P/V_S$ ratios greater than 1.78, similar to the high values common in the adjacent St. Lawrence Platform (where the geometric mean seismic ratio is 1.79). Moreover, and more importantly, the average $V_P$ values for the Grenville province (a region well sampled by refraction profiles) requires low $V_S$ values that are inconsistent with a more mafic composition. Alternative explanations that allow for both high $V_P/V_S$ and reduced $V_P$, $V_S$ include the possibility that Grenville crust contains higher fluid concentrations than that of other tectonic provinces. [[Double checkout less quartzite argument, this implies higher Vp ... but the data does not show it ]]

\begin{table}
  \begin{tabular}{ p{2cm} p{2cm} p{2cm} p{2cm} p{2cm} p{2cm} p{2cm}}
    \hline
    Oxide, wt\% & Smithson [1978] & Rudnick and Fountain [1995] & Christensen and Mooney [1995] & Rudnick and Gao [2003] & This Work: Canada & This Work: Shield\\
    \hline
    $SiO_2$ & 63.0 & 52.6 & 61.7 & 60.6 & 64.3 & 61.9 \\
    $TiO_2$ & 0.7 & 0.8 & 0.9 & 0.7 & 0.6 & 0.7 \\
    $Al_2O_3$ & 15.8 & 16.6 & 14.7 & 15.9 & 15.5 & 15.9 \\
    $Fe_2O_3$ & 2.0 & --- & 1.9 & --- & 1.9 & 2.1 \\
    $FeO$ & 3.4 & 6.6 & 5.1 & 6.7 & 3.3 & 3.9 \\
    $MnO$ & 0.1 & 0.1 & 0.1 & 0.1 & 0.1 & 0.1 \\
    $MgO$ & 2.8 & 4.4 & 3.1 & 4.7 & 2.2 & 2.8 \\
    $CaO$ & 4.6 & 6.4 & 5.7 & 6.4 & 4.3 & 5.1 \\
    $Na_2O$ & 4.0 & 3.2 & 3.6 & 3.1 & 3.6 & 3.6 \\
    $K_2O$ & 2.7 & 1.9 & 2.1 & 1.8 & 2.9 & 2.5 \\
    $H_2O$ & --- & --- & 0.8 & --- & 1.1 & 1.2 \\
    $P_2O_5$ & --- & 0.2 & 0.2 & 0.1 & 0.2 & 0.2 \\
    $CO_2$  & --- & --- & --- & --- & 0.1 & 0.1 \\
    \hline
  \end{tabular}
  \caption{Bulk crustal composition comparison from selected sources. Rudnick and Fountain [1995] and Rudnick and Gao [2003] use a combined $Fe$ quantity and is listed under $FeO$}
\label{table:composition}

@@

Since our average seismic parameter estimates for the entire Canadian crust lie on a line connecting diorite with granite gneiss (\ref{fig:lith}), we may estimate bulk crustal composition in proportion to these lithologies as 51\% diorite and 49\% granite. The lithological estimate for the Canadian Shield is 68\% diorite and 32\% granite. This lithological ratio is applied to the chemical compositions of diorite and granite using chemical averages calculated from 141 diorite and 197 granite samples compiled by Le Maitre [1976] to produce an estimate of bulk crustal composition. The resulting chemical breakdown is provided in table \ref{table:composition} along with previous estimates from global crustal studies. Our analysis reveal a crust with marginally higher average $SiO_2$ composition and lower $Fe$ and $Mg$ content than these previous studies. It is noteworthy that that the two studies involving greatest reliance on seismic attributes, ours and the {\it P}-wave velocity study by Smithson [1978], bear the closest similarity. Figure \ref{fig:composition} displays the major crustal element contributions from the data in table \ref{table:composition} normalized to our average Canadian crustal composition. The gray-shaded region indicates the $\pm 30\%$ range from our reference average Canadian crustal model.  Also included in this figure is the our average crustal composition estimate for the Canadian Shield. Its values closely track the global crustal estimates of Christensen \& Mooney [1995] and Smithson [1978]. These data imply that the extensive platform and orogenic regions of Canada skew the average composition away from the global estimates of previous studies through incorporation of a higher proportion of granite gneiss. In contrast, the Canadian Shield crust aligns more closely with these previous global estimates.

@@

>>>>

The lower energy in the 20-30 km depth window noted in Section \ref{section:discontinuities} for the station profiles given in figure \ref{fig:disc} provides some supporting evidence for seismically distinct upper and lower crustal layers. To better compare results across stations, we normalize each depth profile such that the Moho falls on a common ordinate (Figure \ref{fig:discInterp}, left) and stack the absolute values (Figure \ref{fig:discInterp}, right). The stacked signal has highest amplitude near the surface, where energy is associated with sedimentary layers, and at the Moho. There is a steady and marked decrease in energy over the bottom third of the profile that cannot be ascribed to sidelobe structure associated with the Moho pulse. This observation is ostensibly at odds with the Mooney and Brocher [1987] high reflectivity lower crustal model. Moreover, reflection profiles from {\sc Lithoprobe} transects [Hammer et. al., 2010] frequently display significant reflected energy through the crust with some transects including the Western Superior transect and the Cassiar in the Western Cordillera displaying pronounced reflectivity within the lower crust. We reconcile the teleseismic and active source observations of transparent versus reflective lower crust by considering the scale of vertical heterogeneity to which the methods are sensitive. Teleseismic body waves include both direct and reflected conversions that can originate from layering with velocity gradients over scale lengths of the order of $\lambda/2$ and $\lambda/4$, respectively, where $\lambda$ is the {\it P}-wavelength [Bostock, 1999]. For 1 Hz teleseismic waves, crustal layering with scale lengths of 1000-2000 m should be resolvable. Higher frequency active source data (10-40 Hz) employing {\it P}-wave sources are sensitive to finer scale vertical structure with scales of 75-300 m. We infer then that the vertical scales of lower crustal reflectivity are limited to scale lengths under 100 m yielding a lower crustal structure that is relatively transparent to teleseismic waves but can yield strongly reflective signatures to near-vertical, active-source seismic profiling.  Some authors, for example Mayer et .al. [1996] in their study of the Rhine Graben, have employed the term Conrad discontinuity to to refer to an interface or transition between transparent upper crust and reflective lower crust. It is possible in this context that the higher energy levels between 15 and 20 km on the 5th principal component in Figure 12, is a manifestation of such signals, but the lower energy levels ($<5\%$) in this mode would indicate that Conrad Discontinuity so-defined is at best weakly defined within the Canadian crust. This finding is supported by crustal models characterized by a first order continuously increasing velocity profile such as the gradual transition from a more silicic uppercrust to a more mafic lower crust in shields and platforms observed by Rolf Meisnner [Meisnner, 1986]. It is also supported by the lack of evidence for a consistent mid-crustal reflection boundary in lithoprobe data [Hammer et. al., 2010]. The increased transparency of the lower crust would appear to be better defined and would appear to relate to changes in rheology. Studies on crustal strength [Jackson, 2002] and on brittle and ductile models [Burov and Watts, 2006; Meisner and Rabbel, 2006] as well as the widely noted reduced seismicity of the lower crust indicate changes would support this contention, though see [Deichmann, 1991; Handy and Brun, 2004; Simpson, 1999] for competing evidence.

@@

Seismic parameter estimates for Canada and the Canadian Shield fall, to a first-order approximation, between those of two candidate end-member lithologies: granite geniss and diorite. This intermediate or andesitic composition is in general agreement with other earlier studies of bulk crustal composition e.g., [Rudnick and Fountain, 1995; Christensen and Mooney [1995]. The Slave, Superior and Churchill provinces possess comparable seismic parameters implying similar mean crustal compositions, with the Slave province represented as slightly more felsic in average composition. All three provinces cluster about the average value for the Canadian Shield. The Grenville Province, in contrast, displays markedly lower $V_S$ and higher $V_P/V_S$ implying a significant difference in mean composition. The high $V_P/V_S$ estimate, 1.77, is similar to the averaged results from a previous seismic study [Eaton et. al., 2005] that included stations of the central Grenville Orogen, and might be interpreted to indicate that the Grenville province possesses a mean crustal composition corresponding composition that is more mafic than other Shield provinces. This interpretation is consistent with a tectonic model for the the central Grenville Province which suggests that the mid-crust has been completely excised locally [Eaton et. al., 2005; Rivers, 2008]. However, many stations in the south-east of the province also display $V_P/V_S$ ratios greater than 1.78, similar to the high values common in the adjacent St. Lawrence Platform (where the geometric mean seismic ratio is 1.79). Moreover, and more importantly, the average $V_P$ values for the Grenville province (a region well sampled by refraction profiles) requires low $V_S$ values that are inconsistent with a more mafic composition. Alternative explanations that allow for both high $V_P/V_S$ and reduced $V_P$, $V_S$ include the possibility that Grenville crust contains higher fluid concentrations than that of other tectonic provinces. [[Double checkout less quartzite argument, this implies higher Vp ... but the data does not show it ]]

\begin{table}
  \begin{tabular}{ p{2cm} p{2cm} p{2cm} p{2cm} p{2cm} p{2cm} p{2cm}}
    \hline
    Oxide, wt\% & Pakiser and Robinson [1966] & Smithson [1978] & Christensen and Mooney [1995] & Rudnick and Gao [2003] & This Work: Canada & This Work: Shield\\
    \hline
    $SiO_2$ & 57.9 & 63.0 & 61.7 & 60.6 & 64.3 & 61.9 \\
    $TiO_2$ & 1.2 & 0.7 & 0.9 & 0.7 & 0.6 & 0.7 \\
    $Al_2O_3$ & 15.2 & 15.8 & 14.7 & 15.9 & 15.5 & 15.9 \\
    $Fe_2O_3$ & 2.3 & 2.0 & 1.9 & --- & 1.9 & 2.1 \\
    $FeO$ & 5.5 & 3.4 & 5.1 & 6.7 & 3.3 & 3.9 \\
    $MnO$ & 0.2 & 0.1 & 0.1 & 0.1 & 0.1 & 0.1 \\
    $MgO$ & 5.3 & 2.8 & 3.1 & 4.7 & 2.2 & 2.8 \\
    $CaO$ & 7.1 & 4.6 & 5.7 & 6.4 & 4.3 & 5.1 \\
    $Na_2O$ & 3.0 & 4.0 & 3.6 & 3.1 & 3.6 & 3.6 \\
    $K_2O$ & 2.1 & 2.7 & 2.1 & 1.8 & 2.9 & 2.5 \\
    $P_2O_5$ & 0.3 & --- & 0.2 & 0.1 & 0.2 & 0.2 \\
    \hline
  \end{tabular}
  \caption{Bulk crustal composition comparison from selected sources. Rudnick and Gao [2003] use total $Fe$ expressed as $FeO$.}
\label{table:composition}

@@

Since our average seismic parameter estimates for the entire Canadian crust lie on a line connecting diorite with granite gneiss (\ref{fig:lith}), we may estimate bulk crustal composition in proportion to these lithologies as 51\% diorite and 49\% granite. A similar estimate for the Canadian Shield is 68\% diorite and 32\% granite. This lithological ratio is applied to the chemical compositions of diorite and granite using chemical averages calculated from 141 diorite and 197 granite samples compiled by Le Maitre [1976] to produce an estimate of bulk crustal composition. The resulting chemical breakdown is provided in table \ref{table:composition} along with previous estimates from global crustal studies. Our analysis reveal a crust with marginally higher average $SiO_2$ composition and lower $MgO$ content than these previous studies. Figure \ref{fig:composition} displays the major crustal element contributions from the data in table \ref{table:composition} normalized to our average Canadian crustal composition. The gray-shaded region indicates the $\pm 30\%$ range from our reference average Canadian crustal model.  Also included in this figure is the our average crustal composition estimate for the Canadian Shield. Its values closely track the global crustal estimates of Christensen \& Mooney [1995] and Smithson [1978]. These data imply that the extensive platform and orogenic regions of Canada skew the average composition away from the global estimates of previous studies through incorporation of a higher proportion of granite gneiss. In contrast, the Canadian Shield crust aligns more closely with these previous global estimates.

@@

Meissner, R., (Ed.) (1986) \it{The Continental Crust: A Geophysical Approach}, 426, Academic, San Diego, Calif.
